Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-07-19 paper/paper.tex
WIP

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_fig.png}
\caption{Simulated retinal and tectal signalling molecule expression. Four
receptor types are expressed, and four corresponding ligand types. Receptors
of type $i$ are activated by ligands of type $i$.
%
\textbf{A} Retinal receptor expression, $r_i$. Four receptors are expressed and
displayed here in two dual-colour maps. $r_0$ and $r_1$ are red and blue;
$r_2$ and $r_3$ are cyan and magenta. $r_0$ increases as an exponential
function of $-x$; $r_1$ increases with an exponential function of $-y$. $r_2$
and $r_3$ have the opposite sense. Retinal receptor expression is modelled
with exponential functions because there is ample evidence to support this
form. $N$: nasal, $T$: temporal, $D$: dorsal, $V$: ventral.
%
\textbf{B} Retinal ligand expression, $l_i$, is expressed in complimentary
gradients \citep{hornberger_modulation_1999}.
%
\textbf{C} Tectal ligand expression $L_i$. Tectal ligand expression is shown in a linear form,
as there is less evidence to determine whether ligand expression is
exponential, linear or logarithmic on the tectum.
$L$: lateral, $M$: medial, $C$: caudal, $R$: rostral.
}
\label{f:1}
\end{figure}

Each retinal ganglion cell projects $n=8$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands,
$l_i$, indexed by $i$ and
expressed at levels determined by the cell's location on the retina. The
expression levels of the receptors and ligands vary with respect to the cell's
position on the retinal surface. Each receptor or ligand varies with respect
only to one dimension.  Receptor expression gradients are arranged in
orthogonal pairs with the gradient of $r_0$ being orthogonal to that of
$r_1$. $r_2$, whose gradient is opposite to $r_0$, is orthogonal to
$r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$ opposing
$r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to $l_0$ and
opposing $r_1$.

@@

Figure~\ref{f:1} shows colour maps of receptor and ligand
expression, in this case tectal ligand expression maps are linear.

@@

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, of each
branch, $b$, is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$ and an axon-axon interaction, $\mathbf{I}_b$:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{I}_b
\end{equation}
%
evaluated at time $t$. $m_1$ and $m_2$ are scalar parameters.

@@

@@

First, we describe the behaviour of the model with $m_2 = 0$, which has only
the gradient-based chemoaffinity effect and no axon-axon interaction.

>>>>

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_fig.png}
\caption{Simulated retinal and tectal signalling molecule expression. Four
receptor types are expressed, and four corresponding ligand types. Receptors
of type $i$ are activated by ligands of type $i$.
%
\textbf{A} Retinal receptor expression, $r_i$. Four receptors are expressed and
displayed here in two dual-colour maps. $r_0$ and $r_1$ are red and blue;
$r_2$ and $r_3$ are cyan and magenta. $r_0$ increases as an exponential
function of $-x$; $r_1$ increases with an exponential function of $-y$. $r_2$
and $r_3$ have the opposite sense. Retinal receptor expression is modelled
with exponential functions because there is ample evidence to support this
form. $N$: nasal, $T$: temporal, $D$: dorsal, $V$: ventral.
%
\textbf{B} Retinal ligand expression, $l_i$, is expressed in complimentary
gradients \citep{hornberger_modulation_1999}.
%
\textbf{C} Tectal ligand expression $L_i$. Tectal ligand expression is shown in a linear form,
as there is less evidence to determine whether ligand expression is
exponential, linear or logarithmic on the tectum.
$L$: lateral, $M$: medial, $C$: caudal, $R$: rostral.
}
\label{f:ex}
\end{figure}

Each retinal ganglion cell projects $n$ growth cone agents (referred to
as \emph{branches}) which carry a set of 4 receptors, $r_i$, and 4 ligands,
$l_i$, indexed by $i$ and
expressed at levels determined by the cell's location on the retina. The
expression levels of the receptors and ligands vary with respect to the cell's
position on the retinal surface. Each receptor or ligand varies with respect
only to one dimension.  Receptor expression gradients are arranged in
orthogonal pairs with the gradient of $r_0$ being orthogonal to that of
$r_1$. $r_2$, whose gradient is opposite to $r_0$, is orthogonal to
$r_3$. Ligands are also arranged in orthogonal pairs, with $l_0$ opposing
$r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$ orthogonal to $l_0$ and
opposing $r_1$.

@@

Figure~\ref{f:ex} shows colour maps of receptor and ligand
expression, in this case tectal ligand expression maps are linear. Note that
the expression is discretized, and in simulations, the gradient is computed
numerically from the discretized expression. This makes it straightforward to
perform tectal and retinal `graft' manipulations.

@@

At each timestep in a simulation, the position, $\mathbf{x}_{b,t}$, of each
branch, $b$, is updated according to
%
\begin{equation}
\mathbf{x}_{b,t+1} = \mathbf{x}_{b,t} + \mathbf{M}_{b}
\end{equation}
%
where $\mathbf{M}_{b}$, the movement vector, is the sum of a chemoaffinity
effect, $\mathbf{G}_b$, an axon-axon interaction, $\mathbf{I}_b$ and a `border effect',
$\mathbf{B}_b$, which acts to retain branches within the tectal region:
%
\begin{equation}
\mathbf{M}_{b} = m_1 \mathbf{G}_b + m_2 \mathbf{I}_b + m_3 \mathbf{B}_b
\end{equation}
%
evaluated at time $t$. $m_1$, $m_2$ and $m_3$ are scalar parameters.

@@

\subsection*{Border effect}

We implement a border effect based on gradient following by assuming that
there is some other molecular signal which acts on all branches near the
boundary of the tectal tissue. For a branch with position $(x,y)$, $\mathbf{B}_b$ is
given by:
%
\begin{equation}
B_{b,x} = \begin{cases}
        2r-x    & x<2r \\
        1-2r-x    & x>1-2r
\end{cases}
B_{b,y} = \begin{cases}
        2r-y    & y<2r \\
        1-2r-y    & y>1-2r
\end{cases}
\end{equation}
%
Thus $\mathbf{B}_b$ is equivalent to the action of a repulsive signalling
molecule expressed around the border of the tectum, whose expression increases
quadratically outside the tectum.

@@

First, we describe the behaviour of a model which has only the gradient-based
chemoaffinity effect and no axon-axon interaction or border effect
(i.e.~$m_2=m_3=0$). This model is run with $n=8$ branches per RGC axon. All
receptor-ligand signals induce repulsive interactions (i.e.~gradient
descent). Receptor and ligand expressions are as shown in
Fig.\,\ref{f:ex}. Fig.\,\ref{f:ch}A gives a colour map of retinal ganglion
cell soma locations, with the colour red encoding $x$ position on the
temporal-nasal axis (with red signifying nasal positions). Green encodes $y$
position on the dorso-ventral axis.
%
Fig.\,\ref{f:ch}A+ shows the expected final location of each RGC axon,
according to experimental observations in a `fishnet' plot, which draws lines
between adjacent axons.
%
Fig.\,\ref{f:ch}B shows the position of
400 branches at $t=500$ (in arbitrary time units). Considering each
coordinate axis separately, each branch interacts differentially with two
opposing gradients. Branches which settle nearest the rostral edge of the
tectum interact more strongly with the gradient $L_0$ than with $L_2$
(Fig\,\ref{f:ex}C). A similar pair of gradients separate branches along the
M-L tectal axis. The branches become positioned on a grid due to the
discretization of the tectal ligand gradient (as the number of branches and
tectal elements tends towards infinity, the arangement tends towards a
continuous map---see Fig.\,S1 \textbf{FIXME}).

\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_chemo.png}
\caption{Chemoaffinity model}
\label{f:ch}
\end{figure}

Fig.\,\ref{f:ch}C shows the RGC axon centroids; the mean location of $n$
branches per axon. Because all $n$ branches for each RGC have identical
receptor expression and they experience only the chemoaffinity effect, each
branch attains almost the same final position and so the centroid graph
strongly resembles the graph of all the branches. Fig.\,\ref{f:ch}C also shows
